---
title: "Table 1 and unweighted CIL and AUC"
author: "Yiyao Chen"
date: "10/30/2020"
output:
  bookdown::html_document2:
    toc: true
  html_document:
    df_print: paged
header-includes: \usepackage{natbib}
---

```{r setup, include=FALSE}
library(knitr)
library(ggplot2)
library(ggfortify)
source('/Users/yiyaochen/Binary_SELECT_PLCO/Prepare_ReadIn.R')
```

```{r}
nrow(SELECT)
nrow(PLCO)
```


# Table 1
```{r}
PLCO$cohort <- rep(1, nrow(PLCO))
SELECT$cohort<- rep(0, nrow(SELECT))

Total <- rbind(PLCO[, c('PSA','DRE_binary', 'PSA4flag', 'Age', 'FH', 'Race', 'PB', 'biopsyflag', 'cohort')],
               SELECT[, c('PSA','DRE_binary','PSA4flag', 'Age', 'FH', 'Race', 'PB','biopsyflag', 'cohort')])

Total$agegroups <- Total$Age
Total$agegroups[which(Total$Age <= 60)] <- 1
Total$agegroups[which(Total$Age > 60 & Total$Age <= 70)] <- 2
Total$agegroups[which(Total$Age > 70 & Total$Age<= 80)] <- 3
Total$agegroups[which(Total$Age > 80 )] <- 4

Total$agegroups3 <- Total$Age
Total$agegroups3[which(Total$Age <= 65)] <- 1
Total$agegroups3[which(Total$Age > 65 & Total$Age <= 75)] <- 2
Total$agegroups3[which(Total$Age > 75 )] <- 3



Total$PSAgroups <- Total$PSA
Total$PSAgroups[which(Total$PSA <= 4)] <- 1
Total$PSAgroups[which(Total$PSA > 4 & Total$PSA <= 10)] <- 2
Total$PSAgroups[which(Total$PSA > 10)] <- 3

Total$FH <- as.factor(Total$FH)
Total$Race <- as.factor(Total$Race)
Total$PB <- as.factor(Total$PB)
```

The following table compares the characteristics at first PSA test among PLCO and SELECT participants. P values from Wilcoxon tests for age and PSA or chi-squared tests for categorical risk factors are given. 
```{r}
t_age <- as.data.frame.matrix(table(Total$agegroups, Total$cohort))
t_age$`0` <- paste0(t_age$`0`, ' (', round(t_age$`0`/sum(t_age$`0`)*100, 1), ')')
t_age$`1` <- paste0(t_age$`1`, ' (', round(t_age$`1`/sum(t_age$`1`)*100, 1), ')')
rownames(t_age) <- c('Age ≤ 60','Age in (60, 70]', 'Age in (70, 80]', 'Age > 80')
t_age$p.value <- c(ifelse(wilcox.test(PLCO$Age, SELECT$Age)$p.value < 0.001, '< 0.001', round(wilcox.test(PLCO$Age, SELECT$Age)$p.value, 3)), NA, NA, NA)

t_age3 <- as.data.frame.matrix(table(Total$agegroups3, Total$cohort))
t_age3$`0` <- paste0(t_age3$`0`, ' (', round(t_age3$`0`/sum(t_age3$`0`)*100, 1), ')')
t_age3$`1` <- paste0(t_age3$`1`, ' (', round(t_age3$`1`/sum(t_age3$`1`)*100, 1), ')')
rownames(t_age3) <- c('Age ≤ 65','Age in (65, 75]',  'Age > 75')
t_age3$p.value <- c(ifelse(wilcox.test(PLCO$Age, SELECT$Age)$p.value < 0.001, '< 0.001', round(wilcox.test(PLCO$Age, SELECT$Age)$p.value, 3)), NA, NA)


t_psa <- as.data.frame.matrix(table(Total$PSAgroups, Total$cohort))
t_psa$`0` <- paste0(t_psa$`0`, ' (', round(t_psa$`0`/sum(t_psa$`0`)*100, 1), ')')
t_psa$`1` <- paste0(t_psa$`1`, ' (', round(t_psa$`1`/sum(t_psa$`1`)*100, 1), ')')
rownames(t_psa) <- c('PSA ≤ 4','PSA in (4, 10]', 'PSA > 10')
t_psa$p.value <- c(ifelse(wilcox.test(PLCO$PSA, SELECT$PSA)$p.value < 0.001, '< 0.001', round(wilcox.test(PLCO$PSA, SELECT$PSA)$p.value, 3)), NA, NA)

t2 <- as.data.frame.matrix(table(Total$FH, Total$cohort))
t2$`0` <- paste0(t2$`0`, ' (', format(round(t2$`0`/sum(t2$`0`)*100, 1), nsmall = 1), ')')
t2$`1` <- paste0(t2$`1`, ' (', round(t2$`1`/sum(t2$`1`)*100, 1), ')')
rownames(t2) <- c('No family history','Has family history')
t2$p.value <- c(ifelse(chisq.test(table(Total$FH, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$FH, Total$cohort))$p.value, 3)), NA)
                         
t3 <- as.data.frame.matrix(table(Total$Race, Total$cohort))
t3$`0` <- paste0(t3$`0`, ' (', round(t3$`0`/sum(t3$`0`)*100, 1), ')')
t3$`1` <- paste0(t3$`1`, ' (', round(t3$`1`/sum(t3$`1`)*100, 1), ')')
rownames(t3) <- c('No African ancestry','African ancestry')
t3$p.value <- c(ifelse(chisq.test(table(Total$Race, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$Race, Total$cohort))$p.value, 3)), NA)

t <- as.data.frame.matrix(table(Total$PB, Total$cohort))
t$`0` <- paste0(t$`0`, ' (', round(t$`0`/sum(t$`0`)*100, 1), ')')
t$`1` <- paste0(t$`1`, ' (', round(t$`1`/sum(t$`1`)*100, 1), ')')
rownames(t) <- c('No prior negative biopsy before enrollment','Has prior negative biopsy before enrollment')
t$p.value <- c(ifelse(chisq.test(table(Total$PB, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$PB, Total$cohort))$p.value, 3)), NA)

t4 <- as.data.frame.matrix(table(Total$DRE_binary, Total$cohort))
t4$`0` <- paste0(t4$`0`, ' (', round(t4$`0`/sum(t4$`0`)*100, 1), ')')
t4$`1` <- paste0(t4$`1`, ' (', round(t4$`1`/sum(t4$`1`)*100, 1), ')')
rownames(t4) <- c('DER = Normal','DRE = Abnormal')
t4$p.value <- c(ifelse(chisq.test(table(Total$DRE_binary, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$DRE_binary, Total$cohort))$p.value, 3)), NA)

t5 <- as.data.frame.matrix(table(Total$PSA4flag, Total$cohort))
t5$`0` <- paste0(t5$`0`, ' (', round(t5$`0`/sum(t5$`0`)*100, 1), ')')
t5$`1` <- paste0(t5$`1`, ' (', round(t5$`1`/sum(t5$`1`)*100, 1), ')')
rownames(t5) <- c('PSA <= 4 ng/ml','PSA > 4 ng/ml')
t5$p.value <- c(ifelse(chisq.test(table(Total$PSA4flag, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$PSA4flag, Total$cohort))$p.value, 3)), NA)

t6 <- as.data.frame.matrix(table(Total$biopsyflag, Total$cohort))
t6$`0` <- paste0(t6$`0`, ' (', round(t6$`0`/sum(t6$`0`)*100, 1), ')')
t6$`1` <- paste0(t6$`1`, ' (', round(t6$`1`/sum(t6$`1`)*100, 1), ')')
rownames(t6) <- c('Unverified','Verified')
t6$p.value <- c(ifelse(chisq.test(table(Total$biopsyflag, Total$cohort))$p.value < 0.001, '< 0.001', round(chisq.test(table(Total$biopsyflag, Total$cohort))$p.value, 3)), NA)



t <- rbind(t_age,t_age3, t_psa, t4, t5, t2, t3, t, t6)
colnames(t) <- c(paste0('SELECT (Total = ', nrow(SELECT), ')'), paste0('PLCO (Total = ', nrow(PLCO), ')'), 'P value')
options(knitr.kable.NA = '')
kable(t, format = 'markdown')
```


# Distributions of PSA in SELECT and PLCO
The following plot compares the distribution of PSA in PLCO and SELECT with x-axis shown in log-based-10 scale.
```{r, fig.align = 'center', fig.width = 7, fig.height = 5, warning=FALSE}
dt <- data.frame(PSA = c(PLCO$PSA, 
                          SELECT$PSA),
                 Group = c(rep('PLCO',nrow(PLCO)), rep('SELECT',nrow(SELECT))))

p <- ggplot(dt, aes(x = PSA, fill = as.factor(Group), color = as.factor(Group)))+
  geom_density(alpha = 0.5, size = 1)+
  theme_bw()+
  xlab('PSA')+
  ylab('Density')+
  scale_x_log10(breaks = c(0.01, 0.1, 1, 10, 100, 500, 1138), limits = c(0.01, 1138.5),
                labels = c('0.01', '0.1', '1.0','10.0','100.0','500.0', '1138.0'))+
  scale_fill_manual(values = c("steelblue","red2"),
                    breaks=c("PLCO", "SELECT"),
                    labels = c(paste0("PLCO (N = ", nrow(PLCO), ")"),
                               paste0("SELECT (N = ", nrow(SELECT), ")")))+
  scale_color_manual(values = c("steelblue","red4"),
                    breaks=c("PLCO", "SELECT"),
                    labels = c(paste0("PLCO (N = ", nrow(PLCO), ")"),
                               paste0("SELECT (N = ", nrow(SELECT), ")")))+
  theme(legend.position = c(0.15, 0.9),
        legend.background = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(face="bold"),
        axis.text.x = element_text(face="bold",size=10),
        axis.text.y = element_text(face="bold",size=10),
        axis.title.x = element_text(face="bold",size=10),
        axis.title.y = element_text(face="bold",size=10))
# ggsave(filename="/Users/yiyaochen/Binary_SELECT_PLCO/PLCO_SELECT_PSA_density.png", p, width = 7, height = 5, dpi = 500)
p

```

# Models for cancer risk and being verified in PLCO built on PLCO
```{r}
# model for cancer risk
rg_cancer_plco <- glm(formula = PCa ~ log2PSA +Race + FH+ PB+ log2PSA:DRE_binary,
                      family = binomial(link = "logit"), 
                      data = PLCO[which(PLCO$biopsyflag == 1), ])
round(exp(cbind(coef(rg_cancer_plco), confint(rg_cancer_plco))), 4)

# model for being verified in PLCO
rg_pv_plco <- glm(formula = biopsyflag ~ log2PSA + PSA4flag + DRE_binary + Age + log2PSA:DRE_binary + 
                    PSA4flag:DRE_binary, family = binomial(link = "logit"), data = PLCO)
round(exp(cbind(coef(rg_pv_plco), confint(rg_pv_plco))), 4)
BIC(rg_pv_plco)

rg_pv_plco_no_interact <- glm(formula = biopsyflag ~ log2PSA + PSA4flag + DRE_binary + Age, family = binomial(link = "logit"), data = PLCO)
round(exp(cbind(coef(rg_pv_plco_no_interact), confint(rg_pv_plco_no_interact ))), 4)
BIC(rg_pv_plco_no_interact)
```

# Apply the developed models to SELECT
```{r}
#################################################
SELECT$pred_cancer_plco <- predict(rg_cancer_plco, newdata = SELECT, probability = TRUE,type = "response" )
SELECT$pred_pv <- predict(rg_pv_plco, newdata = SELECT, probability = TRUE,type = "response" )
```


# Unweigted statistics
## CIL
The 95% confidence interval of estimated unweighted CIL is $\widehat{CIL} \pm 1.96 se(\widehat{CIL)}$.
```{r}
SELECT_biopsy <- SELECT[which(SELECT$biopsyflag == 1), ]

CIL <- sum((SELECT_biopsy$pred_cancer_plco - SELECT_biopsy$PCa))/nrow(SELECT_biopsy)
se_CIL <- sqrt(sum((SELECT_biopsy$pred_cancer_plco - SELECT_biopsy$PCa)^2)/(nrow(SELECT_biopsy)^2) - CIL^2/nrow(SELECT_biopsy))

paste0('CIL (95% CI): ', round(CIL, 3), ' (', round(CIL - qnorm(0.975)*se_CIL, 3), ',', round(CIL + qnorm(0.975)*se_CIL, 3), ')')
```

## AUC
The 95% confidence interval for unweighted AUC follows Delong's method.
```{r, warning=FALSE, message = FALSE}
library(pROC)
t1 <- roc(SELECT_biopsy$PCa ~ SELECT_biopsy$pred_cancer_plco,direction="<", ret = 'coords', ci = TRUE)
paste0('AUC (95% CI (Delong)): ', round(t1$auc[1], 3)*100,'%', ' (', round(t1$ci[1], 3)*100,'%', ',', round(t1$ci[3], 3)*100,'%', ')')
```

# Weighted $CIL_{pot}$ and $AUC_{pot}$
```{r, include=FALSE}
auc_boot <- function(TPR, FPR){
  # inputs already sorted, best scores first 
  dFPR <- c(diff(FPR),0)
  dTPR <- c(diff(TPR), 0)
  AUC <- sum(TPR * dFPR) + sum(dTPR * dFPR)/2
  return(AUC)
}

```

```{r}
CIL_pot <- sum((SELECT_biopsy$pred_cancer_plco - SELECT_biopsy$PCa)* SELECT_biopsy$pred_pv)/ sum(SELECT_biopsy$pred_pv)

SELECT_biopsy <-  SELECT_biopsy[order(SELECT_biopsy$pred_cancer_plco, decreasing=TRUE),]
SELECT_biopsy$l1 <- SELECT_biopsy$PCa*SELECT_biopsy$pred_pv
SELECT_biopsy$l2 <- (1-SELECT_biopsy$PCa)*SELECT_biopsy$pred_pv
SELECT_biopsy$TPR_dt <- cumsum(SELECT_biopsy$l1)/sum(SELECT_biopsy$l1)
SELECT_biopsy$FPR_dt <- cumsum(SELECT_biopsy$l2)/sum(SELECT_biopsy$l2)
auc_pot <- auc_boot(SELECT_biopsy$TPR_dt, SELECT_biopsy$FPR_dt)

dt = data.frame(CIL = paste0(round(CIL_pot, 3)),
                  AUC = c(paste0(round(auc_pot, 3)*100, '%')))
  colnames(dt) <- c(paste0('CIL_pot in SELECT (N = ', nrow(SELECT_biopsy), ')'),
                    paste0('AUC_pot in SELECT (N = ', nrow(SELECT_biopsy), ')'))
  rownames(dt) <- c('Weighted')
  kable(dt, format = 'markdown')

    
```